---
title: "Practical Fall 2019 Completed"
author: "Joshua Strauss
date: "10/28/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = "~/Documents/UMich_Bio201_F19/")
```

As with each lab, add "completed" to the Rmd filename; update the name in the author line above; update the location of the working directory as necessary. 

# Load packages
```{r Load packages, eval=TRUE, echo=TRUE, message=FALSE, warning=FALSE, results='hide'}
library(tidyverse)
library(readxl)
library(broom)
library(cowplot)
set.seed(7)
```

# Part 1 - concepts 
Type the answers to questions 1-5 in the space below. 

### Question 1

Given vector v1 below, will it be read as character or numeric? Why does this happen? [1 point]
```{r}
v1 <- c("one", "2", "three", "4")

typeof(v1)
```

<It's a character because of the quotation marks, which signify a character name rather than a numeric value. Also beyond that, there are literal words also in the quotes, so even if the numbers weren't quoted, there'd still have to be those quoted character words.> 

### Question 2

dataframe1 has 323 rows and dataframe2 has 345 rows. After an inner join the resulting data fame has 312 rows. Why does the result have fewer rows? [1 point]


<Inner joins only join matching data and drop non matching data pairs. We know that some data will have N/A fill ins, so when data is inner joined the rows with these N/As are dropped causing the less rows that we see.>

### Question 3

What are the three assumptions for a t-test? Explain what to do if the variance assumption is violated. Explain what to do if the normality assumption is violated. [3 points]

<
1. Large enough data set. This usually means 30 or more data points, in our case usually participants.


Check this def.
2. Equality of Variance. This means the compared datasets are equally varied.
  - If this is violated you can state it within your t-test then still do the t-test, but not ANOVAs, which you would instead use the non-parametric Kruskall-Wallace.
  
3. Normality. The amount the data fits a normal curve/distribution (or bell-shaped).
  If the normality is violated, you cannot use a t-test. Instead use a non-parametric test like the wilcox test.> 

### Question 4

You have a data frame (called butyrate_df) with the columns: "Participant_ID", "Study_week", "Semester", "Supplement_consumed", "Avg_butyrate". If this data frame is piped to each of the functions below, evaluate if the description of the results is true or false. [3 points]
```{r eval=FALSE, include=FALSE}

butyrate_df %>%
  filter(Semester == "Fall2019")
# Result1 = keeps only data from Fall 2019 semester

butyrate_df %>%
  filter(Avg_butyrate <= 47.0)
# Result2 = keeps only values equal to 47 mmol/kg 

butyrate_df %>%
  group_by(Semester) %>%
  summarize(New_column = median(Avg_butyrate))
# Result3 = new_column contains the median of each participant 
```

<1. True, the filter function will keep only the data with the marked response.

2. False, Result2 would actually keep all values with 47mmol/kg and below because the function used was "<=" instead of "==". The former is less than or equals to rather than the latter, equals.

3. False, this will give the median for each semester because you are grouping by semester.> 

### Question 5


The data frame (called breath_df) has the columns: "Participant_ID", "Study_week", "Semester", "Supplement_consumed", "Methane", "Hydrogen", "Carbon_dioxide". This data frame is used for the code below, explain what will happen to any NAs in each of the shown calculations, and how the NAs will effect the underlying data. [2 points]
```{r eval=FALSE, include=FALSE}
# Option 1
breath_df %>%
  filter(!is.na(Methane)) %>% 
  group_by(Participant_ID, Study_week) %>%
  summarize(Avg_methane = mean(Methane))

# Option 2
breath_df %>%
  group_by(Participant_ID, Study_week) %>%
  summarize(Avg_methane = mean(Methane, na.rm = TRUE))
```

<Both options do the same thing, which is remove NAs from the methane data, this allows the calculations to run. Without it, the NAs would cause the means to show up as NA because the NAs will ruin the computers calculations generally. These function's data in both will show up with numeric values for the mean rather than NAs.> 

# Part 2 - tidy data 

### Question 6

Find the data frame matching your unique name in the GitHub repository. Find the breath gas and taxa abundance data frames. Download and import all into RStudio and use these data frames to complete the rest of the practical. Name the dataframe to match your uniqname, name the breath gas data frame "breath", name the taxa abundance data frame "taxa". [1 point]
```{r}
# data import code 
joshuays <- read_delim("~/UMich_Bio201_F19/Practial/Raw_data/joshuays.txt", 
    "\t", escape_double = FALSE, trim_ws = TRUE)

taxa <- read_delim("~/UMich_Bio201_F19/Practial/Raw_data/taxa.txt", 
    "\t", escape_double = FALSE, trim_ws = TRUE)


breath <- read_delim("~/UMich_Bio201_F19/Practial/Raw_data/breath.txt", 
    "\t", escape_double = FALSE, trim_ws = TRUE)

```

Update column names in all files match the course conventions. [1 point]
```{r}
joshuays<-joshuays %>% 
  rename_all(tolower)

taxa<-taxa %>% 
  rename_all(tolower)

breath<-breath %>% 
  rename_all(tolower)

```

View the supplement types in your semester. Are they all entered in the same style? If not update them to match. [1 point]
```{r}
joshuays[["supplement_consumed"]] <- recode(.x=joshuays[["supplement_consumed"]], "Arabinoxylan"="Arabino")

count(joshuays, supplement_consumed)
```

### Question 7

What are the dimensions of each data frame? Write the results as a comment after the function. [1 point]
```{r}
nrow(joshuays)
ncol(joshuays)

nrow(taxa)
ncol(taxa)

nrow(breath)
ncol(breath)

#joshuays dimensions =  445 rows by 9 columns
#taxa dimensions = 46881 rows by 6 columns
#breath dimensions = 2668 rows by 5 columns
```
Comment: joshuays dimensions =  445 rows by 9 columns
taxa dimensions = 46881 rows by 6 columns
breath dimensions = 2668 rows by 5 columns



Select a join that will drop as few data points as possible. Join the data frame containing the participant data with the breath gas data. Name this "joined_df". [1 point]
```{r}
joined_df<- full_join(breath, joshuays)
```

What are the dimensions of the joined data frame? Write the results as a comment after the function. [1 point]
```{r}
nrow(joined_df)
ncol(joined_df)
#joined_df dimensions = 3859 rows x 11 columns

```
joined_df dimensions = 3859 rows x 11 columns


### Question 8

Calculate the total SCFAs. [1 point]
```{r}
joined_df <- joined_df %>% 
  mutate(total_scfa_mmol_kg = acetate_mmol_kg + butyrate_mmol_kg + propionate_mmol_kg)
  


```

Calculate the weekly mean of each SCFA of each participant. Calculate the weekly mean methane and hydrogen levels of each participant. [1 point]
```{r}
joined_df <- joined_df %>% 
  filter(!is.na(butyrate_mmol_kg)) %>%
  filter(!is.na(acetate_mmol_kg)) %>%
 filter(!is.na(total_scfa_mmol_kg)) %>%
  filter(!is.na(propionate_mmol_kg)) %>%
  group_by(participant_id, study_week, supplement_consumed, frequency) %>%
  summarize(mean_butyrate = mean(acetate_mmol_kg),mean_acetate = mean(acetate_mmol_kg), mean_propionate = mean(propionate_mmol_kg), mean_h2 = mean(h2), mean_ch4 = mean(ch4), mean_total_scfa_mmol_kg =mean(total_scfa_mmol_kg))

joined_df
```

What are the dimensions of the data frame with the avearges? Write the results as a comment after the function.
```{r}
nrow(joined_df)
ncol(joined_df)
#Dimensions: 125 rows by 10 columns

```
Dimensions: 125 rows by 10 columns


Join this data frame with the one called "taxa". Name the new data frame "all_data". What are the dimensions of this new data frame? [1 point]
```{r}
all_data<-full_join(joined_df, taxa)
nrow(all_data)
ncol(all_data)
#Answer: Dimensions of all_data = 46920 rows x 14 columns

```
Answer: Dimensions of all_data = 46920 rows x 14 columns

I used an full_join instead of an inner join because I want to keep as much data as possible for later use.

Save all data frames generated to your GitHub page. [1 point]
```{r}
write_delim(joined_df, path = "~/UMich_Bio201_F19/Practial/Curated_data/joined_df.txt", delim = "\t")
write_delim(all_data, path = "~/UMich_Bio201_F19/Practial/Curated_data/all_data.txt", delim = "\t")
write_delim(breath, path = "~/UMich_Bio201_F19/Practial/Curated_data/breath.txt", delim = "\t")
write_delim(taxa, path = "~/UMich_Bio201_F19/Practial/Curated_data/taxa.txt", delim = "\t")
write_delim(joshuays, path = "~/UMich_Bio201_F19/Practial/Curated_data/joshuays.txt", delim = "\t")



```

# Part 3 - statistics & graphics 

Update all axes labels and figure titles so they are informative and easy to read. Upload final plots to GitHub. When specified, conduct the appropriate tests, and write the statistic (p, F, R value) as a comment. 

### Question 9

If you have more than one type of supplement consumed in your semester, filter for one type to answer Question 9. Plot each SCFA with the study week (weeks 1 and 3 only) on the x-axis. Make each study week a different color. Hint: you may have to use facets, or create individual plots and combine them into a multi-panelled figure. [3 points]
```{r}
joined_df <- joined_df %>% 
  filter(supplement_consumed == "Arabino")


plot_ace <- joined_df %>%
  ggplot(aes(x = study_week, 
             y = mean_acetate, color = study_week)) +
    geom_violin() + geom_jitter() +
    labs(title = "Mean Acetate Concentratrion for Each Week",
         x = "Study Week",
         y = "Mean Acetate Concentration mmol/kg") +
    theme(axis.text.x = element_blank(),
          legend.position = "none")

plot_but <- joined_df %>%
  ggplot(aes(x = study_week, 
             y = mean_butyrate, color = study_week)) +
    geom_violin() + geom_jitter() +
    labs(title = "Mean Butyrate Concentratrion for Each Week",
         x = "Study Week",
         y = "Mean Butyrate Concentration mmol/kg") +
    theme(axis.text.x = element_blank(),
          legend.position = "none")


plot_pro <- joined_df %>%
  ggplot(aes(x = study_week, 
             y = mean_propionate, color = study_week)) +
    geom_jitter() + geom_violin() +
    labs(title = "Mean Propionate Concentratrion for Each Week",
         x = "Study Week",
         y = "Mean Propionate Concentration mmol/kg") +
    theme(axis.text.x = element_blank())

SCFA_3_Plot<-plot_grid(plot_ace, plot_but, plot_pro, 
          nrow = 1, ncol = 3)
SCFA_3_Plot

ggplot2::ggsave(filename = "SCFA_3_Plot", "~/UMich_Bio201_F19/Practial/Figures/",
                device = "tiff",
                plot = SCFA_3_Plot,
                dpi = "retina")



```

Based on the data presented in the plot you generated above, identify which statistical test you would need to determine if there was a change in SCFA concentration between weeks 1 and 3 of the study. Check the assumptions appropriate for the test. [3 points]








```{r}
# assumptions

#1 Sample Size
joined_df %>%
  filter(supplement_consumed == "Arabino") %>% 
  group_by(study_week) %>% 
  summarise(sample_size = n())



######## 20 sample size is below the general 30 sample size rule for t-tests.

#2 Normality

scfa_1 <- joined_df %>%
  filter(study_week == "week1")

scfa_3 <- joined_df %>%
  filter(study_week == "week3")

shapiro.test(scfa_1$mean_total_scfa_mmol_kg)

shapiro.test(scfa_3$mean_total_scfa_mmol_kg)  

SCFA_Mean_Normality_Test_Wk1<-ggplot(scfa_1, aes(x=mean_total_scfa_mmol_kg)) +
  geom_histogram() +labs(title = "SCFA Mean Normality Test Wk1")
SCFA_Mean_Normality_Test_Wk1

qqnorm(scfa_1$mean_total_scfa_mmol_kg); qqline(scfa_3$mean_total_scfa_mmol_kg)


SCFA_Mean_Normality_Test_Wk3<-ggplot(scfa_3, aes(x=mean_total_scfa_mmol_kg)) +
  geom_histogram() +labs(title = "SCFA Mean Normality Test Wk2")
SCFA_Mean_Normality_Test_Wk3

qqnorm(scfa_3$mean_total_scfa_mmol_kg); qqline(scfa_3$mean_total_scfa_mmol_kg)

########Both are relatively normal. The first dataset is about .08 instead of .1, but as we've discussed, p-values slightly below .1 are still pretty significant. The second is a high 1.2.


ggplot2::ggsave(filename = "SCFA_Mean_Normality_Test_Wk1", "~/UMich_Bio201_F19/Practial/Figures/",
                device = "tiff",
                plot = SCFA_Mean_Normality_Test_Wk1,
                dpi = "retina")
ggplot2::ggsave(filename = "SCFA_Mean_Normality_Test_Wk3", "~/UMich_Bio201_F19/Practial/Figures/",
                device = "tiff",
                plot = SCFA_Mean_Normality_Test_Wk3,
                dpi = "retina")



#Equal Variance



var.test(x = scfa_1$mean_total_scfa_mmol_kg, 
         y = scfa_3$mean_total_scfa_mmol_kg, 
         alternative = "two.sided")

########## The p-value is large at .11 meaning they are equally varied. F= .47

# test 

t.test(x = scfa_1$mean_total_scfa_mmol_kg, 
         y = scfa_3$mean_total_scfa_mmol_kg,
       alternative = "less", paired = FALSE, var.equal = TRUE)


# result?

# p = 1.12 


```

Based on the results of the test, conclude if there was a difference in SCFAs between weeks. 

<The p-value is high at 1.12. This means the two compared datasets are not different. There is no big change from one to the other.

Additionally, I'd like to explain my reasoning to do a t-test. I chose the test I thought would give th most accurate representation of the truth. The data past all the assumptions, except for the sample size. However, I believe that in this instance 20 is a large enough sample size to work, and since it past the other tests, it seemed more accurate to use the t-test over a non-parametric test.> 

### Question 10

Is there a correlation between the abundace of hydrogen generators and breath hydrogen? Create a plot that shows the relationship and conduct the appropriate statistical test(s). [2 points]
```{r}
# plot
H2_Gen_BrH2_Relationship <- all_data %>%
  filter(category == "hydrogen_gen")

H2_Gen_BrH2_Relationship_Plot <- H2_Gen_BrH2_Relationship %>% 
  ggplot(aes(x = mean_h2,
             y = fract)) +
  geom_point() + 
  geom_smooth(method = "lm",
              se = FALSE) +  
  xlab("Breath Hydrogen Amount") + 
  ylab("Abundance of Hydrogen Generators") +labs(title = "Breath Hydrogen Amount vs. Abundance of Hydrogen Generators")

H2_Gen_BrH2_Relationship_Plot

ggplot2::ggsave(filename = "H2_Gen_BrH2_Relationship_Plot", "~/UMich_Bio201_F19/Practial/Figures/",
                device = "tiff",
                plot = H2_Gen_BrH2_Relationship_Plot,
                dpi = "retina")



# correlation test 
LM1 <- H2_Gen_BrH2_Relationship %>%
  lm(mean_h2 ~ fract, data = .) #test relationship
summary(LM1) #view results 

# result?
#p-value = .9
#Adjusted R-Squared value = -.019
```

What can you conclude from these results?

<The R-Squared adjusted value is a much too small for correlation. The p-value is also a bit high. So, no correlation.> 

Is there a correlation between the abundace of hydrogen consumers and breath methane? Create a plot that shows the relationship and conduct the appropriate statistical test(s). [2 points]
```{r}
# plot
H2_Consumers_CH4_Relationship <- all_data %>%
  filter(category == "hydrogen_consumers")

H2_Consumers_CH4_Relationship_Plot <- H2_Consumers_CH4_Relationship %>% 
  ggplot(aes(x = mean_ch4,
             y = fract)) +
  geom_point() + 
  geom_smooth(method = "lm",
              se = FALSE) + #hides confidence interval around line 
  xlab("Breath Methane Amount") + 
  ylab("Abundance of Hydrogen Consumers") +labs(title = "Breath Methan Amount vs. Abundance of Hydrogen Consumers")

H2_Consumers_CH4_Relationship_Plot

ggplot2::ggsave(filename = "H2_Consumers_CH4_Relationship_Plot", "~/UMich_Bio201_F19/Practial/Figures/",
                device = "tiff",
                plot = H2_Consumers_CH4_Relationship_Plot,
                dpi = "retina")


# correlation test 
LM2 <- H2_Consumers_CH4_Relationship %>%
  lm(mean_ch4 ~ fract, data = .) #test relationship
summary(LM2) #view results 

# result?
#p-value = .0225
#Adjusted R-Squared value = .077
```

What can you conclude from these results?

<There is no correlation. The adjusted R-Squared value is far lower than the .8 it would need for significance.> 

### Extra credit:

Is the abundace of hydrogen producers and consumers related? Create a plot, check assumptions, conduct the appropriate statistical test(s), and interpret the results. [1 point]
```{r}
# plot


H2_Gen_Con <- all_data %>%
  filter(category == "hydrogen_gen" | category == "hydrogen_consumers") %>% 
  ggplot(aes(x = category, 
             y = fract, color = category)) +
   geom_violin() + geom_jitter() +
    labs(title = "Hydrogen Generators by Partcipant",
         x = "Category",
         y = "Abundance of Microbes") 

H2_Gen_Con


# assumptions

#1 Sample Size
H2_Gen_Con_Dataset <- all_data %>%
  filter(category == "hydrogen_gen" | category == "hydrogen_consumers") %>% 
  group_by(category) %>% 
  summarise(sample_size = n())
view(H2_Gen_Con_Dataset)



######## >800 sample size is far above the general 30 sample size rule for t-tests.

#2 Normality

H2_Gen <- all_data %>%
  filter(category == "hydrogen_gen")

H2_Con <- all_data %>%
  filter(category == "hydrogen_consumers")

shapiro.test(H2_Gen$fract)

shapiro.test(H2_Con$fract)  

H2_Generators_Normality_Test<-ggplot(H2_Gen, aes(x=fract)) +
  geom_histogram() +labs(title = "H2 Generators Normality Test")+ xlab("Abundance of Hydrogen Generators")
H2_Generators_Normality_Test
qqnorm(H2_Gen$fract); qqline(H2_Gen$fract)

H2_Consumers_Normality_Test<-ggplot(H2_Con, aes(x=fract)) +
  geom_histogram() +labs(title = "H2 Consumers Normality") + xlab("Abundance of Hydrogen Consumers")
H2_Consumers_Normality_Test
qqnorm(H2_Con$fract); qqline(H2_Con$fract)


ggplot2::ggsave(filename = "H2_Generators_Normality_Test", "~/UMich_Bio201_F19/Practial/Figures/",
                device = "tiff",
                plot = H2_Generators_Normality_Test,
                dpi = "retina")
ggplot2::ggsave(filename = "H2_Consumers_Normality_Test", "~/UMich_Bio201_F19/Practial/Figures/",
                device = "tiff",
                plot = H2_Consumers_Normality_Test,
                dpi = "retina")




########Both are so tiny! Neither are normal. The graphs look skewed to one side like half a normal curve. The normal qq line is exponential.P= <2.2e-16,P= <2.2e-16

#Equal Variance



var.test(x = H2_Gen$fract, 
         y = H2_Con$fract, 
         alternative = "two.sided")

########## Again the value is miniscule. Not equally varied. P= <2.2e-16  F=3.25

# test 

wilcox.test(x = H2_Gen$fract, 
            y = H2_Con$fract, 
            alternative = "less", paired = FALSE)


# result?

# p-value = 1







```

<According to the data, with a p-value of 1. They are definitely the same, unrelated to each other.> 


-----
end